Fast and adaptive cointegration based model for forecasting financial time series

### 2017-12-11 slides/defense-slides.tex
Version para el seminario CCTVal

\author[Paola Arce]{Paola Arce \\ Thesis advisor: Luis Salinas}
\date{December 11, 2017}
\title[Thesis defense]{Fast and adaptive cointegration based model for forecasting financial time series}
\subtitle{Seminario Inform\'atica CCTVal}

@@

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: High Frequency trading
\item Cointegration and the VECM model
\item Challenges and Hypothesis
\item Contributions
\item Experiments
\item Conclusions
\item Future work
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{HFT: High Frequency Trading}
High frequency trading has motivated computer-driven strategies capable of processing large amount of data in short periods of time.
\begin{figure}
\includegraphics[width=0.7\paperwidth]{img/HFTmarket}
\caption{Percentage of HFT trades of total US Equity Trading. Source: TABB group}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{HFT Revenue}
However, the revenues have fallen dramatically mainly because of more regulation, increasing cost of HFT infrastructure and more competitive algorithms.
\begin{figure}
\includegraphics[width=0.7\paperwidth]{img/HFTmarket2}
\caption{Revenue in the US. Source: TABB group}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{High Frequency data}
High frequency time series are commonly:
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Non-stationary
\item Irregularly spaced over time.
%\item Heterogeneity
\item Heteroscedastic
\item Potentially cointegrated
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\includegraphics[width=0.35\paperwidth]{img/forecast}
\end{figure}
\end{column}
\end{columns}
\vspace{5mm}
This imply that classical statistical models can't be used. Therefore, financial time series forecasting is still a challenge.
\end{frame}

\begin{frame}
\frametitle{Stationarity}
\begin{block}{Strong definition}
A {\color{red}strictly stationary time series} $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$ is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
to that of the time shifted set, more precisely:
\begin{equation*}
{\color{blue}
P\{y_{t_1} \leq
c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace }
\end{equation*}
\noindent $\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}$ where $c_1,\dots,c_L$ are constants.
\end{block}
\end{frame}

@@

A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$ 
\end{frame}

@@

\begin{frame}
\frametitle{Irregularly spaced over time}
\begin{figure}
\includegraphics[width=0.8\paperwidth]{img/hft-ticks-zoom.png}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Heteroscedastic}
Conditional heteroskedasticity identifies non-constant volatility.
Volatility gives a measure of uncertainty about future returns. Many factors affect volatility such as: historical prices, trends, monetary policies and news.
\begin{figure}
\includegraphics[width=0.8\paperwidth]{img/vix}
\end{figure}
\end{frame}

\section{Cointegration and the VECM model}
\begin{frame}
\frametitle{Cointegration}
\begin{figure}
\includegraphics[width=0.8\paperwidth]{img/motivation}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Cointegration definition}

\begin{block}{Cointegration}
The relationship between non-stationary time series can be modelled if a {\bf stationary linear combination} is shown to exist.
When this happens it is said they have a {\bf long-run
equilibrium} relationship and the variables are {\color{red} cointegrated}.
\end{block}

@@

\begin{frame}
\frametitle{Online VECM}
Recently, online learning algorithms have been proposed to solve problems with large data sets main because: 
\begin{itemize}
\item Their simplicity
\item Process one instance at a time
\item The hypothesis is updated every new data arrives
\end{itemize} 
\end{frame}

@@

\begin{frame}
\frametitle{VECM(p) matrix form}
If we have $N$ data points, with $N>p$, the VECM matrix form is as follows:
\small
{\color{blue}
\begin{equation*}
\underbrace{
      \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\intercal \\
   \Delta\mathbf{y}_{p+2}^\intercal \\
   \vdots \\
   \Delta\mathbf{y}_N^\intercal
   \end{bmatrix}}_{\mathbf{Y} } =
\arraycolsep=1.4pt
\underbrace{
\begin{pmat}[{....|}]
   \mathbf{y}_p^\intercal \boldsymbol{\beta} & \Delta \mathbf{y}_p^\intercal & \Delta\mathbf{y}_{p-1}^\intercal & \dots 
                    & \Delta\mathbf{y}_2^\intercal & 1 \cr
   \mathbf{y}_{p+1}^\intercal  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\intercal & \Delta\mathbf{y}_p^\intercal & \dots
                       & \Delta\mathbf{y}_3^\intercal & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\intercal  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\intercal & \Delta\mathbf{y}_{N-2}^\intercal & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\intercal & 1 \cr
   \end{pmat}}_{\mathbf{X}}
\underbrace{
   \begin{bmatrix}
   \boldsymbol{\alpha}^\intercal \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\intercal
   \end{bmatrix}
}_{\mathbf{W}}
+
\underbrace{
\begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\intercal \\
   \boldsymbol{\epsilon}_{p+2}^\intercal \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\intercal \\
   \end{bmatrix}
}_{\mathbf{E}}
\end{equation*}}

We can estimate the model parameters $\mathbf{W}$ using OLS:
\begin{equation}
{\color{blue}
\mathbf{Y} = \mathbf{X}\mathbf{W}+\mathbf{E}
}
\end{equation}

@@

\begin{frame}
\begin{algorithm}[H]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\{\mathbf{x}_1,\dots,\mathbf{x}_N \}$: $N$ input vectors \\
$\{\mathbf{y}_1,\dots,\mathbf{y}_N \}$: $N$ targets \\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{f(\mathbf{x}_{L+1}),\dots,f(\mathbf{x}_N) \}$: model predictions \\
\STATE Initialize $\mathbf{A}=\displaystyle \sum_{t=1}^L \mathbf{x}_t
\mathbf{x}_t^\intercal$
and $\mathbf{b}=\displaystyle \sum_{t=1}^L \mathbf{x}_t \mathbf{y}^\intercal_{t}$
\FOR { $t = L+1$ to $N$ }
        \STATE read new $\mathbf{x}_t$
        \STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal-
\mathbf{x}_{t-L-1} \mathbf{x}_{t-L-1}^\intercal  $
        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
\mathbf{A}^{-1}\mathbf{x}_t$
        \STATE Read new $y_t$
        \STATE $\mathbf{b} = \mathbf{b} + y_t \mathbf{x}_t$
\ENDFOR

\end{algorithmic}
\caption{Online Ordinary Least Squares}
\label{alg:SLAAR}
\end{algorithm}
\end{frame}

@@

\begin{frame}
\frametitle{Adaptive VECM algorithm (AVECM)}
AVECM has the following features: 
\begin{enumerate}
\item Uses VECM only considering a sliding windows of size $L$ of data.
\item We proposed to choose $L$ and the number of lags $p$ in order to maximise the percentage of cointegration $PC$ in the near past. This process was done every time that new data was processed. 
\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
\end{enumerate}
\end{frame}

@@

@@

\begin{frame}
\frametitle{Conclusions II}
\begin{itemize}
\item The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds.
\item We showed that VECM computational limitations can be tackled using a an online approach in a distributed environment.
\item Cointegration information can be used as an integration tool to detect arbitrage opportunities.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Future research}
\begin{itemize}
\item To explore the relationship between cointegration and performance for different assets (stocks, other forex rates).
\item To include more explicative variables such as bid-ask spread and change in volume.
\item The online approach for other econometrics models it will be also worthy of  study.% Many of them could be adapted and used with higher frequency data.
\item  In this thesis, the online version of OLS was studied, but the online version of
ridge regression could also be applied to obtain a solution with better
generalisation capabilities.
\end{itemize}
\end{frame}

@@

%Have a slide for each of the following
%9 Problem statement or Hypothesis
%9 “Why is this a Hard Problem?”
%9 Approach or Methodology
%9 “Why is this Innovative?”
%9 Assumptions and Constraints
%9 Initial Results (Promise of Great Things to Come)
%9 Validation Plan
%9 Limitations and Applicability
%9 Expected Contributions
%9 “What’s beyond this thesis?”
%9 Roadmap of Thesis

@@

@@

%\subsection{Model comparison} 
\begin{frame}
\frametitle{Random walk}
The random walk model is defined as:
\begin{equation}
\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
\label{rwmodel}
\end{equation}
The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
\begin{equation}
\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
\end{equation}
\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
\end{frame}
%
%
\begin{frame}
\frametitle{ARIMA}
A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
\begin{equation}
\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
\end{equation}
\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
\end{frame}

>>>>

\author[Paola Arce]{Paola Arce}% \\ Thesis advisor: Luis Salinas}
\date{December 11, 2017}
\title[Thesis defense]{Fast and adaptive cointegration based model for forecasting financial time series}
\subtitle{Seminario Inform\'atica CCTVal}

@@

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: High Frequency Trading
\item Cointegration and the VECM model
\item Challenges and Hypothesis
\item Contributions
\item Limitations and Applicability
\item Conclusions and Future Research
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{HFT: High Frequency Trading}
High Frequency Trading has motivated computer-driven strategies capable of processing large amounts of data in short periods of time.
\begin{figure}
\includegraphics[width=0.7\paperwidth]{img/HFTmarket}
\caption{Percentage of HFT trades of total US Equity Trading. Source: TABB group}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{HFT Revenue}
However, the revenues have fallen dramatically mainly because of more regulation, increasing the cost of HFT infrastructure and more competitive algorithms.
\begin{figure}
\includegraphics[width=0.7\paperwidth]{img/HFTmarket2}
\caption{Revenue in the US. Source: TABB group}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{High Frequency data}
High frequency time series are commonly:
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item Non-stationary
\item Irregularly spaced over time.
%\item Heterogeneity
%\item Heteroscedastic
\item Potentially cointegrated
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
\begin{figure}
\includegraphics[width=0.35\paperwidth]{img/forecast}
\end{figure}
\end{column}
\end{columns}
\vspace{5mm}
This implies that classical statistical models as such can't be used. Therefore, financial time series forecasting is still a challenge.
\end{frame}

\begin{frame}
\frametitle{Stationarity}
\begin{block}{Strong definition}
A {\color{red}strictly stationary time series} $y_t \in \mathbb{R}$, $\forall t \in \mathbb{Z}$, is one for which the probabilistic behaviour of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
to that of the time shifted set, more precisely:
\begin{equation*}
{\color{blue}
P\{y_{t_1} \leq
c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace }
\end{equation*}
\noindent $\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}$, where $c_1,\dots,c_L$ are constants.
\end{block}
\end{frame}

@@

A weakly stationary time series is a process where the  mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
\noindent with $\lambda(s,t) = E[(y_s-\mu)(y_t - \mu)]$ 
\end{frame}

@@

\begin{frame}
\frametitle{Irregularly spaced over time}
\centerline{
\includegraphics[width=\paperwidth,height=0.7\paperheight]{img/hft-ticks-zoom.png}
}
\end{frame}

%\begin{frame}
%\frametitle{Heteroscedastic}
%Conditional Heteroscedasticity identifies non-constant volatility.
%Volatility gives a measure of uncertainty about future returns. Many factors affect volatility such as: historical prices, trends, monetary policies and news.
%\begin{figure}
%\includegraphics[width=0.8\paperwidth]{img/vix}
%\end{figure}
%\end{frame}

\section{Cointegration and the VECM model}

\begin{frame}
\frametitle{Cointegration}
Cointegration reflects the idea of that some set of non-stationary variables cannot wander too far from each other.
\begin{figure}
\includegraphics[width=0.8\paperwidth]{img/motivation}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Cointegration definition}
%The idea of cointegration is that there is an I(1) process, underlying two or more processes.
\begin{block}{Cointegration}
The relationship between non-stationary time series can be modelled if a {\bf stationary linear combination} is shown to exist.
When this happens it is said they have a {\bf long-run
equilibrium} relationship and the variables are {\color{red} cointegrated}.
\end{block}
\end{frame}

@@

\begin{frame}
\frametitle{Online VECM}
Recently, online learning algorithms have been proposed to solve problems with large data sets mainly because of: 
\begin{itemize}
\item Their simplicity
\item They process one instance at a time
\item The hypothesis is updated every time new data arrives
\end{itemize} 
\end{frame}

@@

\begin{frame}
\frametitle{VECM(p) matrix form}
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\mathbf{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}
If we have $N$ data points, with $N>p$, the VECM matrix form is as follows:
\small
{\color{blue}
\begin{equation*}
\underbrace{
      \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\intercal \\
   \Delta\mathbf{y}_{p+2}^\intercal \\
   \vdots \\
   \Delta\mathbf{y}_N^\intercal
   \end{bmatrix}}_{\mathbf{Y} } =
\arraycolsep=1.4pt
\underbrace{
\begin{pmat}[{....|}]
   \mathbf{y}_p^\intercal \boldsymbol{\beta} & \Delta \mathbf{y}_p^\intercal & \Delta\mathbf{y}_{p-1}^\intercal & \dots 
                    & \Delta\mathbf{y}_2^\intercal & 1 \cr
   \mathbf{y}_{p+1}^\intercal  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\intercal & \Delta\mathbf{y}_p^\intercal & \dots
                       & \Delta\mathbf{y}_3^\intercal & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\intercal  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\intercal & \Delta\mathbf{y}_{N-2}^\intercal & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\intercal & 1 \cr
   \end{pmat}}_{\mathbf{X}}
\underbrace{
   \begin{bmatrix}
   \boldsymbol{\alpha}^\intercal \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\intercal
   \end{bmatrix}
}_{\mathbf{W}}
+
\underbrace{
\begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\intercal \\
   \boldsymbol{\epsilon}_{p+2}^\intercal \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\intercal \\
   \end{bmatrix}
}_{\mathbf{E}}
\end{equation*}}

%We can estimate the model parameters $\mathbf{W}$ using OLS:
%\begin{equation}
%{\color{blue}
%\mathbf{Y} = \mathbf{X}\mathbf{W}+\mathbf{E}
%}
%\end{equation}

@@

\begin{frame}
\begin{algorithm}[H]
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\{\mathbf{x}_1,\dots,\mathbf{x}_N \}$: $N$ input vectors \\
$\{\mathbf{y}_1,\dots,\mathbf{y}_N \}$: $N$ targets \\
$L$: sliding window size ($L<N$) \\
\ENSURE  $\,$ \\
$\{f(\mathbf{x}_{L+1}),\dots,f(\mathbf{x}_N) \}$: model predictions \\
\STATE Initialize $\mathbf{A}=\displaystyle \sum_{t=1}^L \mathbf{x}_t
\mathbf{x}_t^\intercal$
and $\mathbf{b}=\displaystyle \sum_{t=1}^L \mathbf{x}_t \mathbf{y}^\intercal_{t}$
\FOR { $t = L+1$ to $N$ }
        \STATE read new $\mathbf{x}_t$
        \STATE $\mathbf{A} = \mathbf{A} + \mathbf{x}_t \mathbf{x}_t^\intercal-
\mathbf{x}_{t-L-1} \mathbf{x}_{t-L-1}^\intercal  $
        \STATE output prediction $f(\mathbf{x}_t) =  \mathbf{b}^\intercal
\mathbf{A}^{-1}\mathbf{x}_t$
        \STATE Read new $\mathbf{y}_t$
        \STATE $\mathbf{b} = \mathbf{b} + \mathbf{x}_t \mathbf{y}^\intercal_{t}$
\ENDFOR
\end{algorithmic}
\caption{Online Ordinary Least Squares}
\label{alg:SLAAR}
\end{algorithm}
\end{frame}

@@

\begin{frame}
\frametitle{Adaptive VECM algorithm (AVECM)}
AVECM has the following features: 
\begin{enumerate}
\item It uses VECM only considering a sliding windows of size $L$ of data.
\item We proposed to choose $L$ and the number of lags $p$ in order to maximise the percentage of cointegration $PC$ in the near past. This process was done every time that new data was processed. 
\item And to use a distributed environment to make this parameters search, since this is the most expensive routine.
\end{enumerate}
\end{frame}

@@

\section{Limitations and Applicability}

\begin{frame}
\frametitle{Limitations and Applicability}
\begin{itemize}
\item Cointegration information can be used as an integration tool to detect arbitrage opportunities.
\item Use with stocks, forex rates and crypto-currency?
\item Applicable to other non-stationary but cointegrated time series
\end{itemize}
\end{frame}

\section{Conclusions and Future Research}

@@

\begin{frame}
\frametitle{Conclusions II}
\begin{itemize}
\item The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds.
\item We showed that VECM computational limitations can be tackled using a an online approach in a distributed environment.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Future research}
\begin{itemize}
\item To add {\bf matrix optimizations} to obtain VECM parameters.
\item To add a regularization parameter to get better generalisation capabilities using {\bf Ridge Regression} and the {\bf Aggregating Algorithm for Regression}.
\item To include {\bf more explicative variables} such as bid-ask spread and change in volume.
\item The online approach for {\bf other econometrics models} it is also worthy of  study.% Many of them could be adapted and used with higher frequency data.
\end{itemize}
\end{frame}

@@

%Have a slide for each of the following
%9 Problem statement or Hypothesis
%9 “Why is this a Hard Problem?”
%9 Approach or Methodology
%9 “Why is this Innovative?”
%9 Assumptions and Constraints
%9 Initial Results (Promise of Great Things to Come)
%9 Validation Plan

%9 Expected Contributions
%9 “What’s beyond this thesis?”
%9 Roadmap of Thesis

@@

\begin{frame}
\frametitle{The order book}
\begin{figure}
\includegraphics[width=0.7\textwidth]{img/orderbook}
\end{figure}
\end{frame}

@@

\begin{frame}
\frametitle{Random walk}
The random walk model is defined as:
\begin{equation}
\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
\label{rwmodel}
\end{equation}
The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
\begin{equation}
\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
\end{equation}
\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
\end{frame}
%
%
\begin{frame}
\frametitle{ARIMA}
A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
\begin{equation}
\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
\end{equation}
\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
\end{frame}
